The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-12 paper/iphas-dr2.tex

Each of the sightlines reveals a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre at $l=180^\circ$ (Fig.~\ref{fig:l180}) 
we find a population dominated by lowly-reddened 
main sequence stars and an apparent absence of reddened giants.
%consistent with the sightline extinction of XX from Schlegel (cite). 
In contrast, closer towards the galactic
centre at $l=45^\circ$ (Fig.~\ref{fig:l45})
we find a wealth of reddened objects
which appear to be separated into a lowly
and a highly reddened component,
perhaps revealing two distinct parts of the Galaxy.
Finally, in one of our earliest sightlines at $l=30^\circ$
we find a very high number of extremely reddened giants
in addition to an unreddened population
of foreground dwarfs.

The number density of stars in the colour-colour
and colour-magnitude space
can be compared against population synthesis models
to create three-dimensional maps of the extinction
across several kpc \citep{Sale2010,Sale2012}.
Such an extinction map based on our catalogue
is to be released in a separate paper
that accompanies this data release (Sale et al., in preparation).

\begin{figure*}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-180-3.pdf} 
        \includegraphics[width=0.5\textwidth]{./plots/cmd-180-3.pdf}
    \end{minipage}
    \caption{Colour-colour and colour-magnitude diagram (left and right panel)
    showing sources flagged as \emph{veryReliable}
    located in an area of one square degree
    centred near the Galactic anti-centre 
    at $(l,b)=(180^\circ,+3^\circ)$.
    The colour-colour diagram shows the
    position of the main sequence (thin solid line),
    giant stars (thick solid line)
    and the reddening track for an A0V-type star (dashed line)
    based on the \citet{Pickles1998} library of empirical spectra.
    The colour-magnitude only show the reddening vector
    along with the unreddened 1~Gyr isochrone due to \citet{Bressan2012},
    which has been placed at an arbitrary distance of 2~kpc for reference.
    This is one of the least reddened sightlines
    in the survey %(Av XX, Schlegel et al. XXXX)
    and hence the observed stellar population appears to be dominated 
    by lowly reddened main sequence stars.}
    \label{fig:l180}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-45-2.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-45-2.pdf}
    \end{minipage}
    \caption{Same as above for $(l,b)=(45^\circ,+2^\circ)$,
    which is one of the highest-density sightlines in the survey,
    revealing a population of stars with a reddening distribution
    that appears to be bi-model.}
    \label{fig:l45}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-30-0.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-30-0.pdf} 
    \end{minipage}
    \caption{Same as above for $(l,b)=(30^\circ,0^\circ)$.
    This is one of the most reddened sightlines in the survey.
    %(XXX, Schlegel et al XXXX).
    }
    \label{fig:l30}
\end{figure*}

>>>>

Each of the sightlines reveals a stellar population
with distinct characteristics.
Firstly, towards the Galactic anti-centre at $l=180^\circ$ (Fig.~\ref{fig:l180}) 
we find a population dominated by lowly-reddened main sequence stars.
This is consistent with the estimated total sightline extinction of $E(B-V)=0.49$
given by \cite{Schlegel1998}, following the correction of \cite{Schlafly2011}.
Looking in more detail we can see that the stellar locus is narrower for M-type dwarfs than for earlier types;
we do not observe M dwarfs experiencing the strongest reddening possible for this sightline.
This implies that extinction is still increasing at distances beyond $\sim 2$~kpc,
where M dwarfs are too faint to be contained in the IPHAS catalogue.
It is also clear that there are no unreddened stars earlier than $\sim$K0 visible,
such stars would be saturated if within a few hundred parsecs.
This therefore suggests that there is a measurable increase in extinction locally.
We also note an relative absence of late type giants which,
due to the relative brevity of the corresponding phase of stellar evolution,
would only account for a small proportion of a volume limited sample.

In contrast, 
closer towards the galactic centre at $l=45^{\circ}$ (Fig.~\ref{fig:l45}) we find a wealth of reddened objects.
In the colour-magnitude diagram it is clear that the stars are split into two distinct groups,
with one significantly redder than the other.
The bluer group is composed of main sequence stars,
with the slope of this group in the colour-magnitude diagram attributable to the significantly increasing extinction.
Meanwhile the redder group is principally composed of red giant stars.
As these stars are intrinsically brighter,
at some given apparent magnitude they will be substantially further away than their main sequence counterparts.
Given that, along this sight-line,
extinction continues to increase with distance,
the red giants we observe will be subject to more reddening than a main sequence star exhibiting the same apparent magnitude.
Therefore, the clear split between the two groups arises due to the combination of increasing extinction and the distinct absolute magnitude ranges of main sequence and giant branch stars.

Finally, in one of our earliest sightlines at $l=30^{\circ}$
we find a very high number of extremely reddened giants
in addition to an unreddened population
of foreground dwarfs.
In contrast to the sightline at $l=45^{\circ}$,
there is no clear group of giant stars visible in the colour-magnitude diagram of Fig.~\ref{fig:l30},
though the red clump stars are manifested as a slight over density which sits roughly 0.4~mags redder than the A0V reddening track.
At $(l,b)=(45^{\circ}, 2^{\circ})$ the giant stars observed exhibit a relatively narrow range of reddenings as the lie beyond most or essentially all of the Galactic dust.
At $(l,b)=(30^{\circ}, 0^{\circ})$ this is not the case:
even at the substantial distances at which we can observe reddened giant stars extinction is continuing to rise and we still find ourselves well within the Galactic dust disc.
It is also apparent that the $(r-i)$ width of both the M dwarfs and early A dwarfs is greater than that in Fig.~\ref{fig:l45}.
This is indicative of a steeper rise in reddening,
both within several hundred parsecs (M dwarfs)
and within a few kpc (early A dwarfs).

Here we have presented only a short qualitative description of the information obtainable from IPHAS colour-colour and colour-magnitude plots.
A more rigorous quantitative analysis of the IPHAS catalogue
can be undertaken to estimate both the stellar density distribution in the Milky Way \citep{Sale2010}
and to create detailed three-dimensional maps of the extinction across several kpc \citep{Sale2009,Sale2012}.
An extinction map based on our catalogue is to be released in a separate paper
that accompanies this data release (Sale et al., in preparation).

\begin{figure*}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-180-3.pdf} 
        \includegraphics[width=0.5\textwidth]{./plots/cmd-180-3.pdf}
    \end{minipage}
    \caption{Colour-colour and colour-magnitude diagram (left and right panel)
    showing sources flagged as \emph{veryReliable}
    located in an area of one square degree
    centred near the Galactic anti-centre 
    at $(l,b)=(180^\circ,+3^\circ)$.
    The colour-colour diagram shows the
    position of the main sequence (thin solid line),
    giant stars (thick solid line)
    and the reddening track for an A0V-type star (dashed line)
    based on the \citet{Pickles1998} library of empirical spectra.
    The colour-magnitude only show the reddening vector
    along with the unreddened 1~Gyr isochrone due to \citet{Bressan2012},
    which has been placed at an arbitrary distance of 2~kpc for reference.
    This is one of the least reddened sightlines
    in the survey %(Av XX, Schlegel et al. XXXX)
    and hence the observed stellar population appears to be dominated 
    by lowly reddened main sequence stars.}
    \label{fig:l180}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-45-2.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-45-2.pdf}
    \end{minipage}
    \caption{Same as above for $(l,b)=(45^\circ,+2^\circ)$,
    which is one of the highest-density sightlines in the survey,
    revealing two groups of stars in colour-magnitude space.}
    \label{fig:l45}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=0.5\textwidth]{./plots/ccd-30-0.pdf}
        \includegraphics[width=0.5\textwidth]{./plots/cmd-30-0.pdf} 
    \end{minipage}
    \caption{Same as above for $(l,b)=(30^\circ,0^\circ)$.
    This is one of the most reddened sightlines in the survey.
    %(XXX, Schlegel et al XXXX).
    }
    \label{fig:l30}
\end{figure*}
